### 2024-03-22 thesis/input_files/data_exploratory.tex
6 march
decision trees

@@

>>>>

- correlation between temperatures and load

@@

This section covers the datasets used in our experiments, attributes and features will be thoroughly described.
Next, we will explain the ETL pipeline that we set up in order ease the workflow of our comparison studies.
Finally, in order to to bettern understand patters within the data, detect outliers and discover interesting patterns we carry out an exploratory data analysis.

\section{GEFCom2014}
The chosen dataset for probabilistic load forecasting, was the data from the GEFCom 2014 competition. The data is freely hosted on Dr. Hong blog \cite{hong2016probabilistic}.
Contestant were asked to provide one month ahead hourly probabilistic forecasts on a rolling basis for 15 consecutive round. To get started, in the first round, organizers provided 69 months of hourly load data and 117 months of hourly temperature data. As the competition task went on, the real observed data of the previous tasks was made available.
The GEFCom2014 Data contains five zip files. We are interested in the probabilistic load forecasting data, which is contained in the GEFCom2014-L\_V2 zip file. Within this subfolder, we a txt with the competition instructions and 15 folders one for each of the consecutive tasks. For each of those we have the prediction of the competition benchmark model and the train file upon which we will fit our model.
\\
Each data from the various task folders are shifted by one months between others.
\\
In order to compare our model performance with the winning entries of the GEFCom2014 competition we will refer to the Provisional\_Leaderboard\_V2 file contained also in the GEFCom2014 Data directory. Notice a quick premise, within this xsls file the authors opted for various scores to compare the entries performance. This choice was motivated by the need 
to give a proportial weight to each task when averaging between them; that is in order to not penalise too much the tasks with higher point spreads. Furthermore, additional factors such as frequency of benchmark outperformance, and robustness and clarity in methodology explaination.
\\
Since we are only interested in the load forecasting track, we can directly compare between models by comparing their pinball scores. Such scores are stored inside the subtab L-score-0/L-score-2 of Provisional\_Leaderboard\_V2.

### 2024-03-22 thesis/input_files/probabilistic.tex
decision trees

% \section{Copula models}
\section{Quantile forest}
Meinshausen \cite{meinshausen2006quantile} extends the idea of random forest \cite{breiman2001random} generalising it, the result is the quantile forest algorithm. Quantile forest allows us to estimate conditional quantiles in a non parametric fashion.
\\
In order to understand this algorithm it is necessary to first cover the theory of decision trees and random forests.
- decision trees
- random forest
- quantile forest
\section{Quantile gradient boosting machine}

@@

\subsection{Kernel herding}
Select the best point forecasting method and create a probabilistic forecast by modelling its model errors with kernel herding (do something like the residual bootstrap ensembles if it is meaningful and possible to implement).

>>>>

% \section{Copula models}
\section{Quantile forest}
Meinshausen \cite{meinshausen2006quantile} extends the idea of random forest \cite{breiman2001random} generalising it, the result is the quantile forest algorithm. Quantile forest allows us to estimate conditional quantiles in a non parametric fashion.
\\
In order to understand this algorithm it is necessary to first cover the theory of decision trees and random forests.
\subsection{Decision trees}
Decision trees methods partition recursively the feature space in a set of binary rectangles and then fit a simple model in each of those partitions (the most straightforward is fitting just a constant).
To get started, we first split the space into two disjoint regions, then we model the response variable by the mean of the observed predicted variables with associated features falling in that specific region. Our goal is selecting the best features and best split point such that to achieve the best generalizing fit.
For an example/visualisation consider the pictures \ref{fig:elements_statistical_learning1} and \ref{fig:elements_statistical_learning2} taken from \cite{hastie2009elements}, where the decision tree algorithm is visualised for a regression problem with two independent variables X2 and X1.
\begin{figure}
    \includegraphics[width=0.5\textwidth]{images/elsii1.png}
    \caption{two-dimensional feature space partitioned by recursive binary splittin}
    \label{fig:elements_statistical_learning1}
  \end{figure}
\\
\begin{figure}
\includegraphics[width=1\textwidth]{images/elsii2.png}
\caption{partition tree and regression model}
\label{fig:elements_statistical_learning2}
\end{figure}
\\
Suppose to partition the feature space into M regions $R_1,\dots,R_M$, then the model reads as follows.
\begin{equation}
    f(x)=\sum\limits_{m=1}^{M}c_m\mathbb{1}(x \in R_m)
\end{equation}
\\
It follows that the function minimising the sum of squares is the one with $\hat{c}_m=mean(y_i|x_i \in R_m)$
Finding the best split in terms of minimum sum of squares is computationally infeasible in practice. Therefore we approximate a solution by approaching the problem in a greedy fashion.
Let $j$ denote the splitting variable, $s$ be the the split point we define the two half planes
\begin{equation}
    R_1(j,s)=\{X|X_j\leq s\} \quad R_2(j,s)=\{X|X_j\>s\}
\end{equation}
Then we search for the $s$ and $j$ that solve
\begin{equation}
    \min_{j,s}\left[\min_{c_1} \sum\limits_{x_i \in R_1(j,s)}(y_i-c_1)^2+\min_{c_2} \sum\limits_{x_i \in R_2(j,s)}(y_i-c_2)^2\right]
\end{equation}
The inner problem is easy, as already pointed out we will have 
\begin{equation}
    \hat{c}_1=mean(y_i|x_i \in R_1) \quad \hat{c}_2=mean(y_i|x_i \in R_2)
\end{equation}
For the outer problem, we scan through all the (j,s) tuples and pick the best pair. 
Next, one or both of these regions from the previous step are split into two more regions; we recurse this process until some stopping condition is triggered (max number of branches, max deth of tree, threshold on the MSE, minimum number of observation in each leaf node).
Notice, this being a greedy algorithm implies that our final solution is guaranteed to be just a local optimum not a global one.
Even though their simplicity, these models have proved themselves to be really powerful, see figure \ref{fig:decision_tree} for a visualisaiton . The most popular among these model is the CART \cite{breiman2017classification} tree, its name comes from the fact that it can handle both classification and regression problems.
\begin{figure}
    \includegraphics[width=\textwidth]{images/decision_tree.png}
    \caption{decision tree regression}
    \label{fig:decision_tree}
    % image generated on coolab to be quick
\end{figure}

\subsection{Random forest}
- quantile forest

\section{Quantile gradient boosting machine}

@@

\subsection{Weather quantiles}
In order to get acquainted with the inner working of the presented methods, this section covers an application explaining practical details and comparing results.
The dataset used is the Melbourne Daily Maximum Temperatures \cite{hdrcde}. It contains the daily maximum temperatures in Melbourne, Australia, from 1981-1990, excluding leap days.
Due to the bimodality of the data, such dataset is commonly used to give a difficult quantile regression problem \cite{hyndman1996estimating}, thus why we chose it. The observed bimodality is that a hot day is likely to be followed by either an equally hot day or one much cooler.

- metrics pinball, picw plot, mae error

\subsection{Kernel herding}
Select the best point forecasting method and create a probabilistic forecast by modelling its model errors with kernel herding (do something like the residual bootstrap ensembles if it is meaningful and possible to implement).
